A Bidirected String Graph Model for Genome Assembly

### 2012-12-06 presentation/paper.tex
paper: bidigraph

\documentclass[10pt]{article}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{comment}
%\usepackage{amsthm}

@@

@@

Every living organism contains a collection of hereditary material called a {\em
genome}.  Typically, each cell of a multicellular organism contains a complete
copy of that organism's genome.  A genome consists of one or more DNA ({\it
deoxyribonucleic acid}) molecules in the form of {\it chromosomes}.  DNA is a
double-stranded molecule, each strand of which is a polymer of {\em
nucleotides}.  A nucleotide, together with the corresponding nucleotide on the
opposite strand, is called a {\it base pair} (abbreviation: {\it bp}).  In DNA,
there are four possible nucleotides: adenine, cytosine, guanine, and thymine,
which are abbreviated as \Base{A}, \Base{C}, \Base{G}, and \Base{T},
respectively.  \Base{A} always pairs with T on the other strand, and \Base{C}
always pairs with \Base{G} on the other strand.  In addition, each strand has a
direction, and the two strands of a DNA molecule are directed opposite one
another.  Furthermore, a DNA molecule may be circular, as is usually the case
for bacterial chromosomes.

@@

{\it DNA sequencing} is the process of determining the base pair sequence of
({\it sequencing}) one or more DNA molecules.  {\it Genome sequencing} is the use
of DNA sequencing to determine the base pair sequence of an entire genome.

@@

Therefore, to sequence an entire genome, which may be millions of
base pairs long, the DNA from the genome must be broken into much smaller
fragments, which are then sequenced.  Following this, the original genome is
computationally reconstructed by a computer program called a {\it genome
assembler} in a process called {\it genome assembly}.  {\it Genome sequencing}
therefore refers to the entire process of sequencing a genome, including both
lab work and computational work, while {\it genome assembly} only refers to the
computational work.\footnote{It usually also takes some technology-specific
computational work to actually prepare the sequences for assembly; for example,
a series of digital photographs of clumps of fluorescently labeled molecules
might need to be interpreted as reads using image-processing algorithms.
However, these programs and/or algorithms generally have already been developed
by the manufacturer of the sequencing machine and work fairly well, and ``genome
assembly'' does not cover these algorithms, although genome assembly algorithms
still may be tuned for specific sequencing technologies.}

@@

The goal of genome assembly is to reconstruct a genome, as defined earlier,
given a set of substrings that may originate from any strand of any DNA molecule
in the genome.  Each substring is called a {\it read} and is the sequence of
some DNA that was sequenced by the sequencing machine.  In general, reads may
contain errors, including {\it substitutions}, where a base pair in the read was
incorrectly read as some other base pair, as well as {\it insertions}, where a
sequence of one or more random base pairs have been inserted into some position
in the read, and {\it deletions}, where a sequence of one or more base pairs
from the original genome is missing from the interior of the read.  Reads are
expected to be uniformly distributed on the genome with a certain coverage $X$
depending on how much data was collected, meaning that each base pair in the
original genome is expected to be included in $X$ different reads on average,
allowing for the reads to be from either strand.  However, some sequencing
technologies produce biased data where the coverage across the genome is much
more variable than that expected by uniform, random sampling.

Reads may contain other information, such as {\it quality scores} or {\it
mate-pair} information, that can provide additional hints to the assembler.

@@

Genomes may contain {\it repeats}, which are subsequences that occur multiple
times in the genome--- possibly adjacent to each other (a {\it tandem repeat}),
but not necessarily.  Repeats may be longer than the length of any individual
read, which makes the genome more difficult to assemble accurately.  However,
real genomes almost always contain repeats of substantial length, so genome
assembly algorithms must be able to handle repeats.

@@

Having four different conditions is required because reads may overlap in a
reverse-complement manner.  An overlap is reverse-complement if and only if one
read in the overlap is used in the original direction and the other read is used
in the reverse-complement direction.  Note that an overlap from condition (4) is
{\it not} a reverse-complement overlap, as it can equivalently be stated as the
last $\LengthVar$ bp of $g$ exactly match the last $\LengthVar$ bp of $f$.
\Figure{fig:overlap_types} shows the three fundamental types of overlaps that
may arise from this definition.

A read may overlap with itself; for example, a read may contain a suffix that
matches its prefix of the same length.  Such an overlap is called a {\it
self-overlap}.

Overlaps may also be {\it inexact}.  An inexact overlap is an overlap where
overlapped regions of the two reads do not match exactly (in forward or
reverse-complement form), but instead differ by a number of base-pair
substitutions, insertions, and/or deletions up to a predetermined upper bound,
such as 5\%.  However, inexact overlaps will not be considered further in this
paper.

@@

An {\it extreme point} occurs when an overlapped region extends to the beginning
or the end of the corresponding read.  An overlap always contains at least two
extreme points.

A {\it containing overlap} is an overlap in which the overlapped region of one
read, the {\it contained read}, is the entire read.  A special case of a
containing overlap is an overlap where the overlapped regions include the
entirety of both reads.  For an exact overlap, this can only happen if the two
reads are either identical or exact reverse-complements of each other.

@@

Overlaps are important because two reads that overlap may come from adjacent
positions on the genome.  Although due to repeats, we cannot be certain that two
overlapping reads actually come from adjacent positions, we {\it can} say for
certain that two reads that do not overlap do not come from adjacent positions
on the genome.  Regardless of this, \Figure{fig:overlap_examples} shows how two
reads that overlap potentially could be combined to form a longer sequence.

@@

A number of faster algorithms for computing overlaps have been designed.  One of
the easier ones to understand is the {\it seed-and-extend} algorithm.  It works
by indexing the reads by short subsequences of a certain length (less than or
equal to than the minimum overlap length) and then only comparing reads that
actually share a subsequence.  Each of these short subsequences is called a {\it
seed}.  Two reads may share a seed in either a forward or reverse-complement
manner.  Given two reads that share a seed, the seed can be extended outwards in
both directions in order to find a maximal matched region, that may be either
forward or reverse-complement, between the two reads.  Then, this maximal
matched region is compared against the definition of an overlap.  In particular,
it must be at least as long as the minimum overlap length, it must have least 2
extreme points over the two reads, and it cannot be the prefix or suffix of both
reads if the match does not extend to the opposite end of at least one of the
reads.  If all of these conditions are met and the same overlap was not already
found through another seed in the same two reads, this overlap is added to the
set of overlaps.

@@

For this project, the seed-and-extend algorithm for computing overlaps was
implemented as the program {\tt compute-overlaps}.  It takes as input a file
that contains a list of reads, as well as a parameter that specifies the minimum
overlap length $L$.  It produces as output (via a data structure serialized to a
file) all exact pairwise overlaps of minimum length $L$ among reads in the input
file, including reverse-complement overlaps and self-overlaps.  The running time
of the algorithm was discussed above.  The memory usage is linear with respect
to the total length of the reads and the number of overlaps, which likely could
be improved.

@@

>>>>

\documentclass[10pt]{article}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{comment}
\usepackage{multicol}
%\usepackage{amsthm}

@@

\newcommand{\KeyTerm}[1]{{\it #1}}
\newcommand{\ProgramName}[1]{{\tt #1}}

@@

Every living organism contains a collection of hereditary material called a
\KeyTerm{genome}.  Typically, each cell of a multicellular organism contains a
complete copy of that organism's genome.  A genome consists of one or more DNA
(\KeyTerm{deoxyribonucleic acid}) molecules in the form of
\KeyTerm{chromosomes}.  DNA is a double-stranded molecule, each strand of which
is a polymer of \KeyTerm{nucleotides}.  A nucleotide, together with the
corresponding nucleotide on the opposite strand, is called a \KeyTerm{base pair}
(abbreviation: \KeyTerm{bp}).  In DNA, there are four possible nucleotides:
adenine, cytosine, guanine, and thymine, which are abbreviated as \Base{A},
\Base{C}, \Base{G}, and \Base{T}, respectively.  \Base{A} always pairs with T on
the other strand, and \Base{C} always pairs with \Base{G} on the other strand.
In addition, each strand has a direction, and the two strands of a DNA molecule
are directed opposite one another.  Furthermore, a DNA molecule may be circular,
as is usually the case for bacterial chromosomes.

@@

\KeyTerm{DNA sequencing} is the process of determining the base pair sequence of
(\KeyTerm{sequencing}) one or more DNA molecules.  \KeyTerm{Genome sequencing}
is the use of DNA sequencing to determine the base pair sequence of an entire
genome.

@@

Therefore, to sequence an entire genome, which may be millions of base pairs
long, the DNA from the genome must be broken into much smaller fragments, which
are then sequenced.  Following this, the original genome is computationally
reconstructed by a computer program called a \KeyTerm{genome assembler} in a
process called \KeyTerm{genome assembly}.  \KeyTerm{Genome sequencing} therefore
refers to the entire process of sequencing a genome, including both lab work and
computational work, while \KeyTerm{genome assembly} only refers to the
computational work.\footnote{It usually also takes some technology-specific
computational work to actually prepare the sequences for assembly; for example,
a series of digital photographs of clumps of fluorescently labeled molecules
might need to be interpreted as reads using image-processing algorithms.
However, these programs and/or algorithms generally have already been developed
by the manufacturer of the sequencing machine and work fairly well, and ``genome
assembly'' does not cover these algorithms, although genome assembly algorithms
still may be tuned for specific sequencing technologies.}

@@

The goal of genome assembly is to reconstruct a genome, as defined earlier,
given a set of substrings that may originate from any strand of any DNA molecule
in the genome.  Each substring is called a \KeyTerm{read} and is the sequence of
some DNA that was sequenced by the sequencing machine.  In general, reads may
contain errors, including \KeyTerm{substitutions}, where a base pair in the read
was incorrectly read as some other base pair, as well as \KeyTerm{insertions},
where a sequence of one or more random base pairs have been inserted into some
position in the read, and \KeyTerm{deletions}, where a sequence of one or more
base pairs from the original genome is missing from the interior of the read.
Reads are expected to be uniformly distributed on the genome with a certain
coverage $X$ depending on how much data was collected, meaning that each base
pair in the original genome is expected to be included in $X$ different reads on
average, allowing for the reads to be from either strand.  However, some
sequencing technologies produce biased data where the coverage across the genome
is much more variable than that expected by uniform, random sampling.

Reads may contain other information, such as \KeyTerm{quality scores} or
\KeyTerm{mate-pair} information, that can provide additional hints to the
assembler.

@@

Genomes may contain \KeyTerm{repeats}, which are subsequences that occur
multiple times in the genome--- possibly adjacent to each other (a
\KeyTerm{tandem repeat}), but not necessarily.  Repeats may be longer than the
length of any individual read, which makes the genome more difficult to assemble
accurately.  However, real genomes almost always contain repeats of substantial
length, so genome assembly algorithms must be able to handle repeats.

@@

Having four different conditions is required because reads may overlap in a
reverse-complement manner.  An overlap is reverse-complement if and only if one
read in the overlap is used in the original direction and the other read is used
in the reverse-complement direction.  Note that an overlap from condition (4) is
{\em not} a reverse-complement overlap, as it can equivalently be stated as the
last $\LengthVar$ bp of $g$ exactly match the last $\LengthVar$ bp of $f$.
\Figure{fig:overlap_types} shows the three fundamental types of overlaps that
may arise from this definition.

A read may overlap with itself; for example, a read may contain a suffix that
matches its prefix of the same length.  Such an overlap is called a
\KeyTerm{self-overlap}.

Overlaps may also be \KeyTerm{inexact}.  An inexact overlap is an overlap where
overlapped regions of the two reads do not match exactly (in forward or
reverse-complement form), but instead differ by a number of base-pair
substitutions, insertions, and/or deletions up to a predetermined upper bound,
such as 5\%.  However, inexact overlaps will not be considered further in this
paper.

@@

An \KeyTerm{extreme point} occurs when an overlapped region extends to the
beginning or the end of the corresponding read.  An overlap always contains at
least two extreme points.

A \KeyTerm{containing overlap} is an overlap in which the overlapped region of
one read, the \KeyTerm{contained read}, is the entire read.  A special case of a
containing overlap is an overlap where the overlapped regions include the
entirety of both reads.  For an exact overlap, this can only happen if the two
reads are either identical or exact reverse-complements of each other.

@@

Overlaps are important because two reads that overlap may come from adjacent
positions on the genome.  Although due to repeats, we cannot be certain that two
overlapping reads actually come from adjacent positions, we {\em can} say for
certain that two reads that do not overlap do not come from adjacent positions
on the genome.  Regardless of this, \Figure{fig:overlap_examples} shows how two
reads that overlap potentially could be combined to form a longer sequence.

@@

A number of faster algorithms for computing overlaps have been designed.  One of
the easier ones to understand is the \KeyTerm{seed-and-extend} algorithm.  It
works by indexing the reads by short subsequences of a certain length (less than
or equal to than the minimum overlap length) and then only comparing reads that
actually share a subsequence.  Each of these short subsequences is called a
\KeyTerm{seed}.  Two reads may share a seed in either a forward or
reverse-complement manner.  Given two reads that share a seed, the seed can be
extended outwards in both directions in order to find a maximal matched region,
that may be either forward or reverse-complement, between the two reads.  Then,
this maximal matched region is compared against the definition of an overlap.
In particular, it must be at least as long as the minimum overlap length, it
must have least 2 extreme points over the two reads, and it cannot be the prefix
or suffix of both reads if the match does not extend to the opposite end of at
least one of the reads.  If all of these conditions are met and the same overlap
was not already found through another seed in the same two reads, this overlap
is added to the set of overlaps.

@@

For this project, the seed-and-extend algorithm for computing overlaps was
implemented as the program \ProgramName{compute-overlaps}.  It takes as input a
file that contains a list of reads, as well as a parameter that specifies the
minimum overlap length $L$.  It produces as output (via a data structure
serialized to a file) all exact pairwise overlaps of minimum length $L$ among
reads in the input file, including reverse-complement overlaps and
self-overlaps.  The running time of the algorithm was discussed above.  The
memory usage is linear with respect to the total length of the reads and the
number of overlaps, which likely could be improved.

@@

\subsection{Overview}

After overlaps have been computed from the reads, the next step is to construct
a graph that models how the reads can be assembled together when the overlaps
are used as evidence.  To do this, define a graph where each vertex corresponds
to a read and each edge corresponds to an overlap.  Two vertices are connected
by an edge if and only if the corresponding reads share an overlap.  In
addition, each edge is labeled with an appropriate DNA sequence, the details of
which will be explained later.

It needs to be decided whether each edge will be directed, undirected, or
something else.  To decide this, we need to consider exactly what we want each
edge to actually represent.  Consider three reads $f$, $g$, and $h$, that
overlap as follows:

\begin{center}
	\begin{tikzpicture}[>=\ReadArrowType]
		\draw[->,style=very thick] (0,   0.6) node[anchor=east] {f} -- (3.0, 0.6);
		\draw[->,style=very thick] (1.0, 0.3) node[anchor=east] {g} -- (4.0, 0.3);
		\draw[<-,style=very thick] (2.0, 0.0) node[anchor=east] {h} -- (5.0, 0.0);
	\end{tikzpicture}
\end{center}

There are 3 overlaps so there will be 3 edges--- one between $f$ and $g$, one
between $g$ and $h$, and one between $f$ and $h$.  As shown, the reads can be
assembled together consistently when used in the order $f,\ g,\ h$; therefore, it
must be possible to walk $f \to g \to h$ in the graph to model this assembly.
In addition, these reads could actually spell out the reverse-complement
sequence instead:

\begin{center}
	\begin{tikzpicture}[>=\ReadArrowType]
		\draw[->,style=very thick] (0,   0.0) node[anchor=east] {h} -- (3.0, 0.0);
		\draw[<-,style=very thick] (1.0, 0.3) node[anchor=east] {g} -- (4.0, 0.3);
		\draw[<-,style=very thick] (2.0, 0.6) node[anchor=east] {f} -- (5.0, 0.6);
	\end{tikzpicture}
\end{center}

Therefore, the walk $h \to g \to f$ must be allowed.  Furthermore, read $f$
directly overlaps with read $h$, so it must also be possible to walk directly $f
\to h$ or $h \to f$.

We might consider a directed graph, but this does not work because the graph
does not allow the reverse-complement walks such as $h \to g \to f$:

\begin{center}
	\begin{tikzpicture}[>=triangle 45]
		\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
		\node (f) at (0, 0) {f};
		\node (g) at (1.7, 0) {g};
		\node (h) at (3.4, 0) {h};
		\draw[->,style=very thick] (f) edge (g);
		\draw[->,style=very thick] (g) edge (h);
		\draw[->,style=very thick,bend left] (f) edge (h);
	\end{tikzpicture}
\end{center}

Similarly, we might consider an undirected graph, but this does not work because
the graph fails to disallow walks such as $g \to h \to f$ that correspond to
assemblies that are not supported by the underlying reads.

\begin{center}
	\begin{tikzpicture}[>=triangle 45]
		\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
		\node (f) at (0, 0) {f};
		\node (g) at (1.7, 0) {g};
		\node (h) at (3.4, 0) {h};
		\draw[-,style=very thick] (f) edge (g);
		\draw[-,style=very thick] (g) edge (h);
		\draw[-,style=very thick,bend left] (f) edge (h);
	\end{tikzpicture}
\end{center}

The fundamental problem with using the directed graph is that at this stage in
the assembly, we don't know in absolute terms which strand a given read is from.
Given two reads that overlap, we can say if they are probably from the same
strand or probably from opposite strands, but we cannot say {\em in absolute
terms} which strand a given read is from.  Therefore, we must allow each read to
be used in either the forward or reverse-complement direction.  This implies
that it needs to be possible to traverse each edge in either direction, since
flipping the order of the reads in an overlap, in effect, implies that they are
reverse-complemented.  The undirected graph satisfies these requirements, but it
fails in another way: it has no guarantee that a given read is used in a
consistent way at a location within any single assembly.

The solution to all these problems is to use a new type of graph called a
\KeyTerm{bidirected graph}:
\begin{center}
	\begin{tikzpicture}[>=triangle 45]
		\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
		\node (f) at (0, 0) {f};
		\node (g) at (1.7, 0) {g};
		\node (h) at (3.4, 0) {h};
		\draw[>->,style=very thick] (f) edge (g);
		\draw[>-<,style=very thick] (g) edge (h);
		\draw[>-<,style=very thick,bend left] (f) edge (h);
	\end{tikzpicture}
\end{center}

\subsection{Bidirected graphs}

A \KeyTerm{bidirected graph} is a graph where a directed head is attached to both
ends of each edge.

There are 3 (or 4, ignoring symmetry) types of \KeyTerm{bidirected edges}.  They
follow directly from the different types of overlaps (see
\Figure{fig:overlap_mapping}).

\begin{figure}
	\begin{center}
		\begin{tabular}{p{1cm}cccc}
			{\footnotesize Overlap} &
			\begin{tikzpicture}[scale=0.6,>=\ReadArrowType]
					\draw[->,style=thick] (2,0) -- (0,0);
					\draw[->,style=thick] (1,0.3) -- (3,0.3);
			\end{tikzpicture}
			&
			\begin{tikzpicture}[scale=0.6,>=\ReadArrowType]
					\draw[->,style=thick] (0,0) -- (2,0);
					\draw[->,style=thick] (3,0.3) -- (1,0.3);
			\end{tikzpicture}
			&
			\begin{tikzpicture}[scale=0.6,>=\ReadArrowType]
					\draw[->,style=thick] (0,0) -- (2,0);
					\draw[->,style=thick] (1,0.3) -- (3,0.3);
			\end{tikzpicture}
			&
			\begin{tikzpicture}[scale=0.6,>=\ReadArrowType]
					\draw[->,style=thick] (2,0) -- (0,0);
					\draw[->,style=thick] (3,0.3) -- (1,0.3);
			\end{tikzpicture}
			\\
			&$\downarrow$ & $\downarrow$& $\downarrow$& $\downarrow$ \\
			{\footnotesize Bidirected edge} &
			\begin{tikzpicture}[scale=0.6,>=triangle 45]
					\draw[<->,style=thick] (0,0) -- (2,0);
			\end{tikzpicture}
			&
			\begin{tikzpicture}[scale=0.6,>=triangle 45]
					\draw[>-<,style=thick] (0,0) -- (2,0);
			\end{tikzpicture}
			&
			\begin{tikzpicture}[scale=0.6,>=triangle 45]
					\draw[>->,style=thick] (0,0) -- (2,0);
			\end{tikzpicture}
			&
			\begin{tikzpicture}[scale=0.6,>=triangle 45]
					\draw[<-<,style=thick] (0,0) -- (2,0);
			\end{tikzpicture}
		\end{tabular}
	\end{center}
	\caption{Mapping from the types of overlaps to the types of bidirected edges.}
	\label{fig:overlap_mapping}
\end{figure}

A \KeyTerm{walk in a bidirected graph} $G$ is a continuous sequence of edges in
$G$ such that if any vertex $v$ is entered through a head inwards, it is exited
through a head outwards (unless it is the end of the path), and vice versa.  An
equivalent formulation is that if we consider a possible walk in a bidirected
graph, such as the possible walk $1 \to 2 \to 3 \to 4$ in the graph shown in
\Figure{fig:bidiwalk_example}, in order to check whether the walk is valid or
not, we must consider every vertex in the possible walk except for the very
first and last vertices.  For each such vertex, we need to verify that each time
that vertex is passed through as part of a walk, the arrow heads adjacent to
that vertex that are used in the relevant part of the walk are pointed in the
same direction {\em relative to the walk}. If there are no such pairs of heads
that point towards each other or point away from each other, the walk is valid.
The red circles in \Figure{fig:bidiwalk_example} show the two pairs of heads
that need to be checked for the possible walk $1 \to 2 \to 3 \to 4$.

Note that this definition is symmetric, which implies that the reverse of a walk
in a bidirected graph is also a valid walk (e.g. the walk $4 \to 3 \to 2 \to 1$
in \Figure{fig:bidiwalk_example}).  A bidirected edge may be traversed in either
direction, possibly even during the same walk.

\begin{figure}
	\begin{center} {\small
		\begin{tikzpicture}[>=triangle 45]
			\tikzstyle{every node} = [circle,fill=blue!20,draw=black];
			\node (1) at (0.00, 0) {1};
			\node (2) at (2.00, 0) {2};
			\node (3) at (4.00, 0) {3};
			\node (4) at (6.00, 0) {4};
			\draw[<->,very thick] (1) edge (2);
			\draw[>-<,very thick] (2) edge (3);
			\draw[<-<,very thick] (3) edge (4);
			\draw[color=red,thick] (2.00, 0) circle (0.8);
			\draw[color=red,thick] (4.00, 0) circle (0.8);
		\end{tikzpicture}
	} \end{center}
	\caption{$1 \to 2 \to 3 \to 4$ and $4 \to 3 \to 2 \to 1$ are both valid walks in
	the bidirected graph below.  (Check circled pairs of heads.)}
	\label{fig:bidiwalk_example}
\end{figure}

\section{Removing contained reads}
